The imprint of global magma oceans on exoplanet demographics

### 2022-12-22 src/tex/ms.tex
separate methods and results

\section{Observational test of the runaway greenhouse transition}
\todo[inline]{Methods and results section of the hypothesis testing part. Introduce the idea of constraining magma ocean frequencies and/or physics with Bayesian stats on exoplanet population; introduce Bioverse; report methods of the hypothesis tests.\\
Caveats: see \citet{Turbet2019} Sect. 3, \citet{Bower2019}}
The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.
This section describes our approach.

@@

%We consider a number of key parameters of transit surveys: The telescope's mirror diameter (or effective diameter for a telescope array) $D$, the total time budget of the survey for observations $t_\mathrm{total}$, the maximum number of observed transits for each target $N_\mathrm{obs, max}$, the slew time between observations $t_\mathrm{slew}$, ... \todo[]{double-check which survey params are needed}
We identify the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a given transit survey:
\begin{itemize}
    \item Prevalent water inventory: The magnitude of radius change at the runaway greenhouse threshold is highly sensitive on water mass fraction. As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern.
    \item Sample size: Since we are looking for a population-level trend, the statistical significance will increase with a larger sample size.
    \item Radius measurement precision: The more precise individual planet radii can be determined, the less smeared out the pattern will be. Good \textit{accuracy} is less important, as long as it does not have a systematic error scaling with stellar irradiance.
    \item Availability and precision of mass measurements: For simple geometric reasons ($\rho \propto R^{-3}$), the expected trend is stronger when measured in bulk density than it is in planet radius space. If transiting planets can be followed up to obtain mass measurements, the statistical significance increases.
\end{itemize}
Besides these main factors, uncertainties in the measured instellations can influence the result, although they are typically small due to the very precise orbital period measurements available for transiting planets.
    This can be different for young host stars when their ages can not be well constrained; in particular the long pre-main sequence phase of M~dwarfs shows a large variation in bolometric luminosity~(see Fig.~\ref{fig:luminosity_tracks}).

\subsubsection{Survey simulations}
\begin{note}
The survey module of Bioverse converts the synthetic planet sample into a set of uncertainty-laden measurements of some parameters of a subset of that sample.
   This task includes selection of the targets, application of any detection biases, and conducting simulated measurements.
   All of the above are specific to the particular survey.

@@

\subsubsection{Detection bias and sample selection}\label{sec:sensitivity}
Not all transiting planets are detectable with the same likelihood and detection biases have an impact on the demographic measurements we are interested in.
A detailed characterization of the detection biases of individual missions would not be justifiable given the uncertainties of the theoretical predictions.
Instead, we derived generic observing limits that reflect the limitations of state-of-the-art transit surveys.

@@

\subsubsection{Definition of the runaway greenhouse hypothesis}
We now turn to assessing the simulated surveys in terms of their ability to detect the runaway greenhouse transition and their constraining power on parameters of this trend.
To do this, we rely on a Bayesian hypothesis testing approach where we quantify the evidence of a hypothesis over another based on the (simulated) data.
For our specific problem, this implies comparing evidences for a demographic imprint of the runaway greenhouse effect to its absence.

@@

    As a null hypothesis, we consider the case where the planetary radius distribution is independent of the instellation,
    \begin{equation}
        h_{\mathrm{null}}(\theta, S) = \theta,
    \end{equation}
    where $\theta$ is the set of parameters defining the radius distribution.
    We further define an alternative hypothesis that describes radius changes due to steam atmospheres and water in the melt.
    As motivated above, this hypothesis takes the form of a step function in instellation $S$, where the step occurs at the outer edge of the runaway greenhouse region.
    Our main observable shall be the average planet radius in the planet population inside and outside this threshold.
    The magma ocean hypothesis is then defined as
\begin{equation}\label{eq:mo_hypo}
    h_{\mathrm{MO}}(\theta, S) =
%        \begin{cases}
%        1 & \text{if } x \in \mathbb{Q}\\
%        0 & \text{if } x \in \mathbb{R}\setminus\mathbb{Q}
%    \end{cases}
        \begin{cases}
        h_{\mathrm{null}} & \text{for } S \leq S_\mathrm{thresh}\\
        \langle R_\mathrm{P}\rangle (f_\mathrm{steam}, f_\mathrm{water}) & \text{for } S > S_\mathrm{thresh}.
    \end{cases}
\end{equation}
    Here, $f_\mathrm{steam}$ and $f_\mathrm{water}$ are predictions from the steam atmosphere and water incorporation models, respectively.
    They are assumed to act additively on the planet radii and thus on their average $\langle R_\mathrm{P}\rangle $.
\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
        \caption{Planets above and below the runaway greenhouse threshold.
        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of instellation. Only the planets marked as transiting are observable. Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, radii are inflated by the amount indicated with gray lines. The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue). This magma ocean hypothesis can be tested agains the null hypothesis $H_0$ (green), where average radii are independent of instellation.}
        \label{fig:HnullHmo}
    \end{centering}
\end{figure}
    Figure~\ref{fig:HnullHmo} shows the null and alternative hypotheses as a function of instellation.
    The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.

@@

\subsubsection{Hypothesis tests}
We are now ready to assess the relative plausibility of $h_{\mathrm{MO}}$ and $h_{\mathrm{null}}$ given the synthetic data we have generated, assigning equal a~priori~ probabilities to these models.
This is done by comparing the Bayesian evidence $\mathcal{Z}$ of the models, which we estimate with the nested sampling~\citep{Skilling2004} algorithm \emph{dynesty}~\citep{Speagle2020}.
Our criterion to reject the null hypothesis is
\begin{equation}
\Delta \ln \mathcal{Z}  = \ln \mathcal{Z} - \ln \mathcal{Z}_\mathrm{null}  > 3.
\end{equation}
The chosen threshold roughly corresponds to the $p = 0.05$ threshold for statistical significance popular in frequentist approaches.\todo{CITATION}

@@

\subsection{Signature of the magma ocean hypothesis}

\subsection{Testability of the magma ocean hypothesis with the PLATO transit survey}
PLATO (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars, a goal to be achieved via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
The PLATO team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission (Matuszewski, priv. comm.).
In the FGK sample of the mission, a minimum of 500 Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
We base our assessment of PLATO's diagnostic power to test the magma ocean hypothesis on these estimates.

@@

\section{Discussion}\label{sec:discussion}
\todo[inline]{\textbf{Potential discussion points: } what is the fraction of Kepler planets w/ runaway GH.}
\subsection{Expected imprint of runaway greenhouse phases on exoplanet demographics}

\subsection{Detectability of the magma ocean demographic feature}

>>>>

\section{Observational test of the runaway greenhouse transition}
\todo[inline]{Methods section of the hypothesis testing part. Introduce the idea of constraining magma ocean frequencies and/or physics with Bayesian stats on exoplanet population; introduce Bioverse; report methods of the hypothesis tests.\\
Caveats: see \citet{Turbet2019} Sect. 3, \citet{Bower2019}}
The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.
This section describes our approach.

@@

\subsection{Survey simulations}
\begin{note}
The survey module of Bioverse converts the synthetic planet sample into a set of uncertainty-laden measurements of some parameters of a subset of that sample.
   This task includes selection of the targets, application of any detection biases, and conducting simulated measurements.
   All of the above are specific to the particular survey.

@@

\subsection{Detection bias and sample selection}\label{sec:sensitivity}
Not all transiting planets are detectable with the same likelihood and detection biases have an impact on the demographic measurements we are interested in.
A detailed characterization of the detection biases of individual missions would not be justifiable given the uncertainties of the theoretical predictions.
Instead, we derived generic observing limits that reflect the limitations of state-of-the-art transit surveys.

@@

\subsection{Definition of the runaway greenhouse hypothesis}
We now turn to assessing the simulated surveys in terms of their ability to detect the runaway greenhouse transition and their constraining power on parameters of this trend.
To do this, we rely on a Bayesian hypothesis testing approach where we quantify the evidence of a hypothesis over another based on the (simulated) data.
For our specific problem, this implies comparing evidences for a demographic imprint of the runaway greenhouse effect to its absence.

@@

    As a null hypothesis, we consider the case where the planetary radius distribution is independent of the instellation,
    \begin{equation}
        h_{\mathrm{null}}(\theta, S) = \theta,
    \end{equation}
    where $\theta$ is the set of parameters defining the radius distribution.
    We further define an alternative hypothesis that describes radius changes due to steam atmospheres and water in the melt.
    As motivated above, this hypothesis takes the form of a step function in instellation $S$, where the step occurs at the outer edge of the runaway greenhouse region.
    Our main observable shall be the average planet radius in the planet population inside and outside this threshold.
    The magma ocean hypothesis is then defined as
\begin{equation}\label{eq:mo_hypo}
    h_{\mathrm{MO}}(\theta, S) =
%        \begin{cases}
%        1 & \text{if } x \in \mathbb{Q}\\
%        0 & \text{if } x \in \mathbb{R}\setminus\mathbb{Q}
%    \end{cases}
        \begin{cases}
        h_{\mathrm{null}} & \text{for } S \leq S_\mathrm{thresh}\\
        \langle R_\mathrm{P}\rangle (f_\mathrm{steam}, f_\mathrm{water}) & \text{for } S > S_\mathrm{thresh}.
    \end{cases}
\end{equation}
    Here, $f_\mathrm{steam}$ and $f_\mathrm{water}$ are predictions from the steam atmosphere and water incorporation models, respectively.
    They are assumed to act additively on the planet radii and thus on their average $\langle R_\mathrm{P}\rangle $.
\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
        \caption{Planets above and below the runaway greenhouse threshold.
        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of instellation. Only the planets marked as transiting are observable. Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, radii are inflated by the amount indicated with gray lines. The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue). This magma ocean hypothesis can be tested against the null hypothesis $H_0$ (green), where average radii are independent of instellation.}
        \label{fig:HnullHmo}
    \end{centering}
\end{figure}
    Figure~\ref{fig:HnullHmo} shows the null and alternative hypotheses as a function of instellation.
    The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.

@@

\subsection{Hypothesis tests}
We are now ready to assess the relative plausibility of $h_{\mathrm{MO}}$ and $h_{\mathrm{null}}$ given the synthetic data we have generated, assigning equal a~priori~ probabilities to these models.
This is done by comparing the Bayesian evidence $\mathcal{Z}$ of the models, which we estimate with the nested sampling~\citep{Skilling2004} algorithm \emph{dynesty}~\citep{Speagle2020}.
Our criterion to reject the null hypothesis is
\begin{equation}
\Delta \ln \mathcal{Z}  = \ln \mathcal{Z} - \ln \mathcal{Z}_\mathrm{null}  > 3.
\end{equation}
The chosen threshold roughly corresponds to the $p = 0.05$ threshold for statistical significance popular in frequentist approaches.\todo{CITATION}

@@

\section{Results}
\subsection{Signature of the magma ocean hypothesis}

%We consider a number of key parameters of transit surveys: The telescope's mirror diameter (or effective diameter for a telescope array) $D$, the total time budget of the survey for observations $t_\mathrm{total}$, the maximum number of observed transits for each target $N_\mathrm{obs, max}$, the slew time between observations $t_\mathrm{slew}$, ... \todo[]{double-check which survey params are needed}
We identified the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a given transit survey:
\begin{itemize}
    \item Prevalent water inventory: The magnitude of radius change at the runaway greenhouse threshold is highly sensitive on water mass fraction. As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern.
    \item Sample size: Since we are looking for a population-level trend, the statistical significance will increase with a larger sample size.
    \item Radius measurement precision: The more precise individual planet radii can be determined, the less smeared out the pattern will be. Good \textit{accuracy} is less important, as long as it does not have a systematic error scaling with stellar irradiance.
    \item Availability and precision of mass measurements: For simple geometric reasons ($\rho \propto R^{-3}$), the expected trend is stronger when measured in bulk density than it is in planet radius space. If transiting planets can be followed up to obtain mass measurements, the statistical significance increases.
\end{itemize}
Besides these main factors, uncertainties in the measured instellations can influence the result, although they are typically small due to the very precise orbital period measurements available for transiting planets.
This can be different for young host stars when their ages can not be well constrained; in particular the long pre-main sequence phase of M~dwarfs shows a large variation in bolometric luminosity~(see Fig.~\ref{fig:luminosity_tracks}).

PLATO (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars, a goal to be achieved via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
The PLATO team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission (Matuszewski, priv. comm.).
In the FGK sample of the mission, a minimum of 500 Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
We base our assessment of PLATO's diagnostic power to test the magma ocean hypothesis on these estimates.

@@

\section{Discussion}\label{sec:discussion}
\todo[inline]{\textbf{Potential discussion points: } what is the fraction of Kepler planets w/ runaway GH.}
\subsection{Detectability of the runaway greenhouse threshold}

\subsubsection{False positive scenarios}
\todo{what other processes could be confused with a magma ocean signal?}
\begin{note}
    Global magma oceans are not the only physical mechanism that may cause a decrease in transit radii for a subset of planets.
    Other causes of shrinking planet sizes have been put forward, the most widely discussed ones being atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
    Both processes are being traded as potentially sculpting the observed radius bimodality of small, close-in exoplanets from the  mission~\citep{Fulton2017,VanEylen2018}.
    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
    This region is distinct from the one affected by global magma oceans, since... \todo{...elaborate why they cannot be confused}

\subsubsection{OPTIONAL...: Atmospheric signatures}
\todo[inline]{discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (Tim's talk)}
